---
title: "Project title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

    I found the data set on kaggle.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    This data, however, was originally collected by the University of California at Irvine. It was collected as a synthesis of data from the Hungarian Institute of Cardiology and the Cleveland Clinic foundation.

-   Write a brief description of the observations.

    1.  id: (Unique id for each patient)

    2.  age: (Age of the patient in years)

    3.  dataset: (place of study)

    4.  sex: (Male/Female)

    5.  cp: chest pain type (\[typical angina, atypical angina, non-anginal, asymptomatic\])

    6.  trestbps: resting blood pressure (resting blood pressure (in mm Hg on admission to the hospital))

    7.  chol: (serum cholesterol in mg/dl)

    8.  fbs: (if fasting blood sugar \> 120 mg/dl)

    9.  restecg: (resting electrocardiographic results)\
        \-- Values: \[normal, stt abnormality, lv hypertrophy\]

    10. thalach: maximum heart rate achieved

    11. exang: exercise-induced angina (True/ False)

    12. oldpeak: ST depression induced by exercise relative to rest

    13. slope: the slope of the peak exercise ST segment

    14. ca: number of major vessels (0-3) colored by fluoroscopy

    15. thal: \[normal; fixed defect; reversible defect\]

    16. num: the predicted attribute 0 means no heart disease, and 1-4 is the stage.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    How does heart disease presence and magnitude vary based on the age, sex, cholesterol, blood sugar, and blood pressure? Furthermore, what variables are the best predictor of whether or not an individual will get heart disease?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    Heart disease affects hundreds of thousands of people across the world. It can come in many different shapes and forms, such as coronary artery disease, heart failure, and heart valve disorders. I hypothesize that the older an individual is, the higher cholesterol, blood sugar, and blood pressure and individual has, the more likely they are to develop a heart disease of large magnitude. I also believe that the age, sex, cholesterol, blood sugar, blood pressure, and the number of major vessels are the best predictors of whether or not an individual will develop heart disease.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    1.  age: quantitative

    2.  dataset: categorical

    3.  sex: categorical

    4.  cp: categorical

    5.  trestbps: quantitative

    6.  chol: quantitative

    7.  fbs: categorical

    8.  restecg: categorical

    9.  thalach: quantitative

    10. exang: categorical

    11. oldpeak: quantitative

    12. slope: categorical

    13. ca: categorical

    14. thal: categorical

    15. num: categorical

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
heart_disease <- read_csv("data/heart_disease_uci.csv")
```

# Data 2

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

# add code here
```

# Data 3

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

# add code here
```
